---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r echo=FALSE}
knitr::opts_chunk$set(message=FALSE, warning = FALSE)
```

## Loading and preprocessing the data
#### Load the required libraries

```{r LoadLibraries}
library(data.table)
library(ggplot2)
library(Hmisc)
library(plyr)
library(knitr)
```

#### Unzip the Activity data file and load into a Data Table

```{r LoadData}
unzip("./activity.zip")

actDT <- data.table::as.data.table(x = read.csv("./activity.csv",header = TRUE, 
                                                sep=",",stringsAsFactors = FALSE))
actDT$date <- data.table::as.IDate(actDT$date,"%Y-%m-%d")
```

## Q1-What is mean total number of steps taken per day?

#### Calculate the total number of steps taken per day,mean and median values

```{r TotalSteps}
stepsDay <- actDT[, round(sum(steps)), by=date]
names(stepsDay) <- c("date", "StepCount")

meanStepsDay <- actDT[, round(mean(steps, na.rm = TRUE)), by=date]
medianStepsDay <- actDT[, round(median(steps, na.rm = TRUE)), by=date]
names(meanStepsDay) = c("date", "mean")
names(medianStepsDay)= c("date", "median")
actMeanMedian <- merge(meanStepsDay,medianStepsDay)

```

#### Plot a histogram of the total number of steps taken per day

```{r HistogramTotalSteps}
ggplot(stepsDay, aes(stepsDay$date, stepsDay$StepCount), na.action=na.omit) + 
  geom_histogram(na.rm = TRUE, stat = "identity") +
    theme(plot.title=element_text(face=2, hjust = 0, color="blue")) +
  labs(title = "Day wise Total Steps", x="Date", y="Total Steps") 
```

#### Report Mean and Median of the total number of steps taken per day

```{r PrintMeanMedian}
knitr::kable(actMeanMedian, caption = "Mean and Median of the total number of steps taken per day", format = "markdown")
```

## Q2-What is the average daily activity pattern?

#### Compute averaged steps taken by interval and the maximum number of steps taken

```{r MaxStepsInterval}
stepsInterval <- actDT[, mean(steps, na.rm = TRUE), by=interval]
names(stepsInterval) =c("Interval", "AverageSteps")

stepsIntervalMax <- stepsInterval[stepsInterval[, AverageSteps==max(AverageSteps, na.rm = TRUE)]]
```

#### Set margins; plot the daily activity pattern and mark the interval when maximum number of steps taken

```{r PlotActivityPattern}
par(mar=c(4,4,3,1), oma=c(1,0,0,0), xpd=FALSE)

with(stepsInterval, plot(Interval, AverageSteps, type="l", col="blue"))
title(main="Average Daily Activity Pattern", col="red", font=2)
points(stepsIntervalMax[,1], stepsIntervalMax[,2], pch=8, col="red")
abline(v=stepsIntervalMax[,1], lty=1, lwd=2, col="red")
axis(side=1, label=TRUE, at=stepsIntervalMax[,1], lwd=2, font = 2)

mtext("08:35-08:40 interval recorded Maximum step count across all days",side=1, 
      outer=TRUE, font = 2, col="blue")
```

## Q3-Imputing missing values
#### Compute and print total number of missing values

```{r PrintMissingValues}
print(paste("Total Number of Missing Values in Dataset is: ", sum(is.na(actDT))))
```

#### Imputing strategy is to replace NAs with the mean of the step count; Add a new column after imputing NAs

```{r ImputeNAs}
actDTImputed <- plyr::mutate(actDT, stepsImputed=Hmisc::impute(actDT$steps,fun=mean))
actDTImputed$date <- as.IDate(actDTImputed$date,"%Y-%m-%d")
```

#### Create a new dataset after imputing missing values for further processing

```{r NewDataset}
actDTImputed_New <- as.data.table(actDTImputed[,c(2,3,4)])
actDTImputed_New$stepsImputed <- as.numeric(actDTImputed_New$stepsImputed)
```

#### Compute day-wise total step count, mean, median and step count

```{r ImputedTotalSteps}
stepsDayImputed <- actDTImputed_New[, round(sum(stepsImputed)), by=date]
names(stepsDayImputed) <- c("date", "StepCount")

meanStepsDayImputed <- actDTImputed_New[, round(mean(stepsImputed, na.rm = TRUE)), by=date]
medianStepsDayImputed <- actDTImputed_New[, round(median(stepsImputed, na.rm = TRUE)), by=date]
names(meanStepsDayImputed) = c("date", "mean")
names(medianStepsDayImputed)= c("date", "median")
actMeanMedianImputed <- merge(meanStepsDayImputed,medianStepsDayImputed)

```

#### Histogram for total number of steps taken each day after imputation

```{r HistTotalStepsImputed}
ggplot(stepsDayImputed, aes(stepsDayImputed$date, stepsDayImputed$StepCount), na.action=na.omit) + 
  geom_histogram(na.rm = TRUE, stat = "identity") +
  theme(plot.title=element_text(face=2, hjust = 0, color="blue")) +
  labs(title = "Day wise Total Steps - After Imputation", x="Date", y="Total Steps") 
```

#### Mean/Median Step count per each day after imputation

```{r MeanMedianImputed}
knitr::kable(actMeanMedianImputed, caption="Mean/Median Step count per each day after imputation", format = "markdown")
```

#### Compare Mean, Median and Total Stepcount values, before and after imputation

```{r CompMeanMedianStepCount}
compMeanMedian <- merge(actMeanMedian,actMeanMedianImputed)
compMeanMedian <- cbind.data.frame(compMeanMedian,stepsDay[,2],stepsDayImputed[,2])
names(compMeanMedian) = c("date","Mean-BI","Median-BI","Mean-AI","Median-AI", "TotalSteps-BI","TotalSteps-AI")
```

#### Comparison table

```{r compMeanMedian}
knitr::kable(compMeanMedian[,c(1,2,4,3,5,6,7)], caption = "Comparison Table", format = "markdown")
```

##### We see that there is no impact on the mean, median and total step count values by imputation except for imputed rows.


## Q4-Are there differences in activity patterns between weekdays and weekends?

#### Add new factor variable "day" for weekend/weekday

```{r Dayfactor}
actDTImputed_New <- as.data.table(mutate(actDTImputed_New, day=factor(ifelse(weekdays(actDTImputed_New$date, abbreviate = TRUE) %nin% 
                                            c("Sat","Sun"),"Weekday","Weekend"))))

```

#### Compute average number of steps taken, averaged across weekday/weekend days

```{r AverageStepsImputed}
stepsIntImputed <- actDTImputed_New[, mean(stepsImputed), by="interval,day"]
names(stepsIntImputed) =c("Interval","day", "AverageStepsImputed")
```

#### Make a panel plot showing Activity Pattern on weekday/weekend days

```{r PanelPlotActivityPattern}
ggplot(stepsIntImputed, aes(Interval,AverageStepsImputed)) + 
  geom_line(col="blue")+
  facet_wrap(.~day, nrow=2, ncol=1) +
  theme(plot.title=element_text(face=2, hjust = 0, color="blue")) +
  labs(title = "Average Daily Activity Pattern - After Imputation", x="Interval", y="Avg Steps")
  
```
